Supermartingales and Termination Analysis of Statistical PCF

### 2020-09-09 probabilisticProgrammingMartingales.tex
Minor: some terms -> skeleton; Proof of Theorem 1

Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term has a skeleton $Sk(M)$ and each skeleton can be converted to a term $S[r]$ given a vector of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. 
\lo{Skeleton is defined in Sec.~\ref{sec:sampling semantics}.}
The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, \changed[lo]{and $M \to M[\skeletonPlaceholder / P]$, in case $M \mid P = \tsample$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$.
(By abuse of notation, we use the same metavariables $M, N$, etc.~for skeletons as for terms.)}
For example, we have 
\(
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
\to
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
\to
\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
\to
\skeletonPlaceholder
\)

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence. 
\akr{I've yet to go through and replace references to terms with references to skeletons, but it should be reasonably obvious where the substitution should take place. 
This is potentially even more confusing though because I'm now implicitly identifying terms, skeletons, reduction sequences and skeletal reduction sequences.} 
Reduction sequences are used rather than reachable \changed[lo]{skeletons} because if the same \changed[lo]{skeleton} is reached twice, different samples may be needed, \lo{Give a simple example illustrating this.} but they will be discussed as though they were terms, by abuse of notation.

@@

The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable \changed[lo]{skeletons}. 
If $(N,P) \sim^* (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either \changed[lo]{both placeholders}, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$. 
\changed[lo]{Formally $L_s(M) := \{[(X, P)]_{\sim^\ast_M} : X \mid P = \tsample\}$.}
The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}

@@

\begin{lemma} \label{downAcrossUp}
If $(N_1,P_1) \sim^* (N_2,P_2)$, there is some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $P'$ in both of them such that $(N_1,P_1) \sim_p^* (N'_1, P') \sim_c^* (N'_2, P') \sim_p^* (N_2, P_2)$.
\end{lemma}
\begin{proof}
The $\sim$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,P_A) \sim_\downarrow (B,P_B)$ if either $A \to B$ and $(A,P_A) \sim_p (B,P_B)$ or $(A,P_A) \sim_c (B,P_B)$, and similarly $(A,P_A) \sim_\uparrow (B,P_B)$ if either $B \to A$ and $(A,P_A) \sim_p (B,P_B)$ or $(A,P_A) \sim_c (B,P_B)$. The sequence of $\sim$ relations linking $N_1$ to $N_2$ can then be expressed as an alternating sequence of $\sim_\downarrow^*$ and $\sim_\uparrow^*$s, starting with $\sim_\downarrow^*$ and ending with $\sim_\uparrow^*$, and the proof proceeds by induction on the number of these alternations.

@@

Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
\begin{align*}
& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(Sk(M),P)}/P]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$}
\end{align*}

@@

\paragraph{}
The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
A \emph{reduction strategy} starting from a closed term $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.

@@

\begin{theorem} \label{thm:AstEquivalence}
A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
\end{theorem}
\begin{proof}
\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}

For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length. The distributions are equal for each length, by induction, as follows.

@@

For any term $N$ which is not a value, there is a unique environment $E$ and redex $R$ such that $N = E[R]$. The position of the hole in $E$ is equal to $\cbv(N)$, so that $N|\cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of environments. If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$. The distribution of next terms, conditional on the previous term, in the case that that previous term reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.

In the other case that $R = \tsample$, the first (term) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$. If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample. For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent. The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$. Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the term by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$. In either case, the distribution of the next term conditional on the previous term, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.

In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal, any by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.

The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal, therefore so is the probability of having reached a value after $n$ steps, therefore so is its limit, the probability of termination, therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
\end{proof}

>>>>

Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has \changed[lo]{(possibly more than one occurrence of the same)} placeholder $\skeletonPlaceholder$, so that each term has a skeleton $S = Sk(M)$ and each skeleton can be converted to a term $S[\overline r]$ given a $\overline r$ vector of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. 
\lo{Need to resolve skeleton here with earlier definition in Sec.~\ref{sec:sampling semantics}.}
The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with \changed[lo]{$\tif{\skeletonPlaceholder < 0}{A_1}{A_2} \to A_i$ nondeterministically; and $M \to M[\skeletonPlaceholder / P]$, in case $M \mid P = \tsample$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$.
(By abuse of notation, we use the same metavariables $M, N$, etc.~for skeletons as for terms.
It should be clear from the context which is meant.)}
For example, we have 
\(
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
\to
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
\to
\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
\to
\skeletonPlaceholder
\)

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence. 
\akr{I've yet to go through and replace references to terms with references to skeletons, but it should be reasonably obvious where the substitution should take place. 
This is potentially even more confusing though because I'm now implicitly identifying terms, skeletons, reduction sequences and skeletal reduction sequences.} 
Reduction sequences of skeletons are used rather than reachable \changed[lo]{skeletons} because if the same \changed[lo]{skeleton} is reached twice, different samples may be needed, \lo{Give a simple example illustrating this.} but they will be discussed as though they were \changed[lo]{skeletons}, by abuse of notation.

@@

The reflexive transitive closure of this relation $\sim$, written $\sim^*_M$, is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*_M$, and each equivalence class can be considered as the same position as it may occur across multiple reachable \changed[lo]{skeletons}. 
If $(N,P) \sim^*_M (O,Q)$, then $N \mid P$ and $O \mid Q$ both have the same shape (i.e.~they're either \changed[lo]{both placeholders}, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$. 
\changed[lo]{Formally $L_s(M) := \{[(X, P)]_{\sim^\ast_M} : X \mid P = \tsample\}$.}
The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined (see e.g.~\citep[Cor.~2.7.3]{AshDD00}).}
\changed[lo]{(To avoid notational clutter, we will often elide the subscript from $\sim^\ast_M$ whenever the starting skeleton $M$ of the reduction sequences is clear form the context.)}

@@

\begin{lemma} \label{downAcrossUp}
If $(N_1,P_1) \sim^* (N_2,P_2)$, there are some descendants $N'_1, N'_2$ of $N_1$ resp.~$N_2$, and a position $P'$ in both of them such that $(N_1,P_1) \sim_p^* (N'_1, P') \sim_c^* (N'_2, P') \sim_p^* (N_2, P_2)$.
\end{lemma}
\begin{proof}
The $\sim$ relation can be split into $\sim_\downarrow \cup \sim_\uparrow$, where $(A,P_A) \sim_\downarrow (B,P_B)$ if either $A \to B$ and $(A,P_A) \sim_p (B,P_B)$ or $(A,P_A) \sim_c (B,P_B)$, and similarly $(A,P_A) \sim_\uparrow (B,P_B)$ if either $B \to A$ and $(A,P_A) \sim_p (B,P_B)$ or $(A,P_A) \sim_c (B,P_B)$. The sequence of $\sim$ relations linking $N_1$ to $N_2$ can then be expressed as an alternating sequence of $\sim_\downarrow^*$ and $\sim_\uparrow^*$s, starting with $\sim_\downarrow^*$ and ending with $\sim_\uparrow^*$, and the proof proceeds by induction on the number of these alternations.

@@

Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
\begin{align*}
& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(Sk(M),P)}/P]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$}
\end{align*}
\lo{$M$ and $N$ above are skeleton. So we should have $N = Sk\big(M[\underline{s(Sk(M),P)}/P]\big)$.}

@@

\paragraph{}
The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
A \emph{reduction strategy} starting from a closed skeleton $M$ is a partial function $f$ from $Rch(M)$ to positions, such that for any reachable skeleton $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.

@@

\begin{theorem} \label{thm:AstEquivalence}
A closed skeleton $M$ is AST with respect to $\cbv$ iff it is AST.
\end{theorem}
\begin{proof}
\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}

For both $\red$ \lo{We have defined the reduction relation $\to$ on skeletons, but not (yet) $\red$.} and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length. 
The distributions are equal for each length, by induction, as follows.

@@

For any skeleton $N$ which is not a value, there is a unique \changed[lo]{context} $E$ and redex $R$ such that $N = E[R]$. 
The position of the hole in $E$ is equal to $\cbv(N)$, so that $N \mid \cbv(N) = R$, as the cases in the definition of $\cbv$ match the cases in the definition of \changed[lo]{context}. 
If $R \neq \tsample$, let $R'$ be the result of reducing $R$ (which is equal in both versions of the semantics), then $E[R'] = N[R' / \cbv(N)]$ therefore $\red(N,s) = (E[R'],s)$ and $(N,s) \Rightarrow_{\cbv} (E[R'],s)$. 
The distribution of next skeletons, conditional on the previous skeleton, in the case that that previous skeleton reduces deterministically, is therefore equal for $\red$ and $\Rightarrow_{\cbv}$.

In the other case that $R = \tsample$, the first (skeleton) part of $\red(N,s)$ is $E[\underline{\pi_h(s)}]$. \lo{This should be $Sk(E[\underline{\pi_h(s)}])$.} 
If $k$ is the number of samples already taken in this reduction sequence, this is equivalent to $E[\underline{s_0(k)}]$, where $s_0$ is the initial sample. \lo{Should introduce a name to replace sample here.}
For any sufficiently small neighbourhood of $N$ (i.e.~containing only reduction sequences with the same skeletons), the probability of reaching this neighbourhood is independent of $s(k)$, because it depends only on the samples taken so far, and the samples are independent. 
The distribution of $s(k)$ conditional on reaching $N$ is therefore the uniform distribution on $I$. 
Similarly for the other semantics, if the distribution of remaining samples after $n$ steps is independent of the skeleton by Lemma \ref{lem:independentSamples}, therefore the distribution of $s(Sk(N),\cbv(N))$ conditional on $N$ is the uniform distribution on $I$. 
In either case, the distribution of the next skeleton conditional on the previous skeleton, in the case that it reduces randomly, is equal to the image under $r \mapsto E[\underline r]$ of the uniform distribution on $I$.

In any case, the distribution of reduction sequences after $n+1$ steps, conditional on the reduction sequence after $n$ steps, as defined by $\Rightarrow_{\cbv}$ and by $\red$, is equal; 
and by the inductive hypothesis the distributions after $n$ steps are equal, therefore by integrating over the reduction sequence up to step $n$, the distributions on reduction sequences up to step $n+1$ are equal.

The distributions on reduction sequences of any finite length as defined by $\Rightarrow_{\cbv}$ and $\red$ are therefore equal; 
therefore so is the probability of having reached a value after $n$ steps; 
therefore so is its limit, the probability of termination; 
therefore the probability of termination is 1 iff the probability of termination with respect to $\cbv$ is 1.
\end{proof}
